Iterative Monomorphisation

### 2024-05-02 iterative_monomorphisation.tex
filled table and improved text

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) more efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause. Generating an excessive number of new clauses can flood the downstream prover. The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the most explosive part of the algorithm. If \(S\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated. Recall that the total number of type argument tuples that are used to generate \(S\) is doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

The monomorphisation algorithm is paramatrised by many bounds, the first step of the evaluation process is to find appropriate values for these bounds. The second step compares the performance of Zipperposition with and without monomorphisation. Finally, the third step tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.

The problems used for these tests are taken from the TPTP library. The TPTP library contains 1765 problems in the TF1 and TH1 grammars, they are the first-order and higher-order problems with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems including real numbers or importing axioms including real numbers. In total our testing suite contains 1754 polymorphic problems. For the first step, we chose as a measure of success the number of problems solved by that instance in under 30 seconds each with a single thread. This setup mirrors the setting of interactive provers in which an instance of an automatic prover will typically be called for similar durations. As a precaution against overfitting, all tests of this phase are made on 500 problems randomly taken out of the 1754 total problems.

@@

Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options with which Zipperposition will be instantiated. Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems. The best performing configuration became the basis upon which we tested the different monomorphisation options. Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.%[TODO recheck this]

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds, however, this is not feasible. Instead, we group closely related bounds together and test combinations of values for bounds in these groups. Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group. This method requires an initial default set of bounds and reasonable ranges of values for each bound. We chose these values based on the results of preliminary evaluations.

Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration. The limit of newly generated type arguments per clause is determined by three components which form a natural group of bounds. Our results show that generating no new monomorphic type arguments seems to be the best approach. It is possible to monorphise formuale without generating monomorphic type arguments because non-monomorphic type argument generation can generate substitutions which instantiate one or more type variables and it is these substitutions which are used to monomorphise formulae.

@@

Table~\ref{nmon_ty_args} fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments. The bound values are lower because in preliminary evaluation we found non-monomorphic type arguments to be explosive. The results confirm that it is non-monomorphic argument generation which drives the creation of useful non-monomorphic clauses. This is indicatied by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Additionally, we observe that removing all bounds considerably reduces performance. %TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with a few extra problems that are easily monomorphised)

@@

The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related as the construction of monomorphising substitutions makes use of all available substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions based on which iteration they were generated in, earlier substitutions are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the specific choice of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
\begin{table}[th]
\caption{Evaluation of bounds for substitution generation}
\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & &&&& \multicolumn{4}{c}{ordering} \\
   & \multicolumn{4}{c}{age} & \multicolumn{4}{c}{random} & \multicolumn{4}{c}{segregation}\\
   \cmidrule(l){2-13}
   & &&&& \multicolumn{4}{c}{mono subst}\\
   \multirow{1}{5.5em}{subst cap} & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 \\
   \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
   50         & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
   100        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
   200        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 171 & 170 & 171\\
   \(\infty\) & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\

@@

Table \ref{pb_size} groups together the bounds related to the size of the output problem to be passed to the E prover. The absolute cap refers to the maximum total number of formulae passed to E. The multiplier instead limits the total number of newly generated formulae based on the problem's initial number of formuale. The performance of the E prover is highly dependent on the number of formulae it is given. To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.

@@

Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the unmodified operation of Zipperposition against the number of iteration of the monomorphisation algorithm. A higher loop number will lead to longer monomorphisation time, we therefore expect better performance whenever loop number and monomorphisation time have coherent values.

>>>>

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) more efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause. Generating an excessive number of new clauses can flood the downstream prover. The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the most explosive part of the algorithm. If \(S\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated. Recall that the total number of type argument tuples that are used to generate \(S\) can be doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

The monomorphisation algorithm is paramatrised by many bounds, the first part of the evaluation process is to find appropriate values for these bounds. The second part compares the performance of Zipperposition with and without monomorphisation. The third part tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.

The test problems are taken from the TPTP library. The TPTP library contains 1765 problems in the TF1 and TH1 grammars, they are the first-order and higher-order problems with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems including real numbers or importing axioms including real numbers. In total our testing suite contains 1754 polymorphic problems. We chose as a measure of success the number of problems that could each be solved by an instance of a prover in under 30 seconds with a single thread. This setup mirrors the setting of interactive provers in which an instance of an automatic prover will typically be called for similar durations.

@@

As a precaution against overfitting, all tests related to parameter optimisation were made on 500 randomly chosen problems out of 1754. Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options with which Zipperposition will be instantiated. Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems without using monomorphisation. The best performing configuration became the basis upon which we tested the different monomorphisation options. Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.%[TODO recheck this]

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds however, this is not feasible. Instead, we group closely related bounds together and test combinations of values for bounds in these groups. Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group. If several values for a given bound solve the same number of problems, the values solving the greater number of problems with monomorphisation will be selected, if these numbers are also equal, the most constraining bounds will be selected. This method requires an initial default set of bounds and reasonable ranges of values for each bound. We determined these values based on the results of preliminary evaluations.

Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration. The limit of newly generated type arguments per clause is determined by three components which form a natural group of bounds. Our results show that generating no new monomorphic type arguments seems to be the best approach. It is possible to monorphise formuale without generating monomorphic type arguments because non-monomorphic type argument generation can generate substitutions which instantiate one or more type variables and it is these substitutions which are used to monomorphise formulae. This table also illustrates the impact that the choice of bounds can have on performance. The choice values of only three bounds can result in an increase of up to 30\% in the number of problems solved.

@@

Table~\ref{nmon_ty_args} fills the same role as table~\ref{mono_ty_args} except it is for bounds limiting the number of new non-monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations. These results confirm that non-monomorphic argument generation drive the creation of useful non-monomorphic clauses. This is indicatied by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems largely unaffected by bounds' values when within a certain range but drops off significantly beyond this range.
%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)

@@

The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
\begin{table}[th]
\caption{Evaluation of bounds for substitution generation}
\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & &&&& \multicolumn{4}{c}{ordering} \\
   & \multicolumn{4}{c}{age} & \multicolumn{4}{c}{random} & \multicolumn{4}{c}{segregation}\\
   \cmidrule(l){2-13}
   & &&&& \multicolumn{4}{c}{mono subst}\\
   \multirow{1}{5.5em}{subst cap} & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 \\
   \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
   %2          & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
   5          & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 &\bf{171}&170&171\\
   50         & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
   100        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
   %200        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 &\bf{171}&170&171\\
   \(\infty\) & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\

@@

Table \ref{pb_size} groups together the bounds related to the size of the output problem to be passed to the E prover. The absolute cap is the maximum number of formulae passed to E. The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae. The performance of the E prover is highly dependent on the number of formulae it is given. To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem. We expect degraded performance when E is given many formulae but little time and few formulae but a lot of time.

@@

For larger problems, the monomorphisation algorithm may timeout despite the bounds. In these cases neither Zipperposition nor E will have had a chance to attempt to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
